---
title: "Cyclistic case study"
author: "Adam Sosnowski"
date: '2022-02-24'
output:
  html_document: 
    theme: cerulean
    toc: TRUE
    toc_float: TRUE
---  

```{r CSS, message=FALSE, warning=FALSE, include=FALSE}
# Code below is a cascading style sheet for HTML webpage rendering
```
<style type="text/css">
h1 {color:#0070C0;}
h2 {color:#ffbd4a;}
h3 {color:#0070C0;}
</style>

```{r Logo, echo=FALSE}
htmltools::img(src = knitr::image_uri("Cyclistic_logo.png"), 
               alt = 'logo', 
               style = 'position:absolute; height:225px; width:225px; top:0; right:0; padding:30px;')
```

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;  
&nbsp;  

## INTRODUCTION

Welcome to the Cyclistic bike-share analysis case study!  In this case study, I am a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago.

&nbsp; 

### Stakeholders
* Lily Moreno - director of marketing and my manager
* Cyclistic marketing analytics team
* Cyclistic executive team

&nbsp; 

### About the company
In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.  

&nbsp;  

Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members.  

&nbsp;  

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, Moreno believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.  

&nbsp;  

Moreno has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends.  

&nbsp;  

### Business task for my analysis:
Answer the question: *“In what ways do members and casual riders use Cyclistic bikes differently?”*

&nbsp;  
&nbsp;  

## PREPARE DATA  

&nbsp;  

### Data sources
I will use Cyclistic’s historical trip data to analyze and identify trends. Data is from April 2019 thru March 2020.
The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement) This data is public.  

I will import necessary packages and the data from four quarters stored in four separate files. I will use R language to prepare R Markdown report, documenting all the steps of my data analysis.

&nbsp;  
&nbsp;  

## PROCESS DATA

&nbsp;  

### Install required packages

- `tidyverse` for data import and wrangling
- `lubridate` for date functions
- `ggplot2` for visualization
- `plotly`
- `htmltools`
- `scales`


```{r Packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(plotly)
library(htmltools)
library(scales)
```

```{r Working Directory, echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
getwd() #displays your working directory
setwd('C:/Users/Adam/Desktop/Data_Analytics_Portfolio/Cyclistic_Case_Study') #sets your working directory
```

&nbsp;  
&nbsp;  

### Upload datasets
Uploading csv files.

```{r Datasets Upload, message = FALSE, warning = FALSE}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

&nbsp;  

### Compare column names
Comparing column names in each of the files.
While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r Compare Column Names, message = FALSE, warning = FALSE}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

&nbsp;  
&nbsp;  

### Rename columns 
Renaming columns to make them consistent with q1_2020 (as this will be the going-forward table design for Cyclistic)

```{r Rename Columns, message = FALSE, warning = FALSE}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))
```

&nbsp;  
&nbsp;  

### Inspect the dataframes
Inspecting the dataframes and looking for inconsistencies.

```{r Inspect Dataframes Structure, message = FALSE, warning = FALSE}
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

&nbsp;  
&nbsp;  

### Convert field types
Converting `ride_id` and `rideable_type` to character so that they can stack correctly


```{r Convert Field Types, message = FALSE, warning = FALSE}
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

&nbsp;  
&nbsp;  

### Review column names again

```{r Review Column Names, message = FALSE, warning = FALSE}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

&nbsp;  
&nbsp;  

### Combine data frames into one
Stack individual quarter's data frames into one big data frame

```{r Combine Data Frames, message = FALSE, warning = FALSE}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

&nbsp;  
&nbsp;  

### Remove unnecessary fields
Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r Remove Unnecessary Fields, message = FALSE, warning = FALSE}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

```

&nbsp;  
&nbsp;  

## CLEAN UP DATA
Cleaning up and adding data to prepare for analysis

&nbsp;  

### Inspect combined data frame
Inspect the new table that has been created

```{r Inspect Combined Data Frame, message = FALSE, warning = FALSE}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame   
tail(all_trips)  #See the last 6 rows of data frame
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

&nbsp;  
&nbsp;  

### Fixing the issues
There are a few problems which will need to be fixed:

  1. In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for   casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
  
  2. The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
  
  3. We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
  
  4. There are some rides where tripduration shows up as negative, including several hundred rides where Cyclistic took bikes out of circulation for Quality Control reasons. We will want to delete these rides.  
    

&nbsp;  
&nbsp;  

**Begin by seeing how many observations fall under each usertype**

```{r Review Usertypes, message = FALSE, warning = FALSE}
table(all_trips$member_casual)
```

&nbsp;  
&nbsp;  

**Reassign to the desired values (we will go with the current 2020 labels)**

```{r Convert Usertypes, message = FALSE, warning = FALSE}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
```

&nbsp;  
&nbsp;  

**Check to make sure the proper number of observations were reassigned**

```{r, message = FALSE, warning = FALSE}
table(all_trips$member_casual)
```

&nbsp;  
&nbsp;  

**Add columns that list the date, month, day, year and day of week of each ride.  **

This will allow us to aggregate ride data for each month, day, or year. 
Before completing these operations we could only aggregate at the ride level.  

```{r Add New Columns, message = FALSE, warning = FALSE}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

&nbsp;  
&nbsp;  

**Add a `ride_length` calculation to all_trips (in seconds)**

```{r, message = FALSE, warning = FALSE}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

&nbsp;  
&nbsp;  

**Inspect the structure of the columns**

```{r, message = FALSE, warning = FALSE}
str(all_trips)
```

&nbsp;  
&nbsp;  

**Convert `ride_length` from Factor to numeric so we can run calculations on the data**

```{r, message = FALSE, warning = FALSE}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

&nbsp;  
&nbsp;  

**Remove "bad" data**  
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Cyclistic or ride_length was negative.  
We will create a new version of the dataframe (v2) since data is being removed.  

```{r, message = FALSE, warning = FALSE}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

&nbsp;  
&nbsp;  

## ANALYSIS

&nbsp;  

**Descriptive analysis on `ride_length` (all figures in seconds)**  

```{r, message = FALSE, warning = FALSE}
mean(all_trips_v2$ride_length)   #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length)    #longest ride
min(all_trips_v2$ride_length)    #shortest ride

```

&nbsp;  
&nbsp;  

**Condensing the four lines above to one line using summary() on the specific attribute**  

```{r, message = FALSE, warning = FALSE}
summary(all_trips_v2$ride_length)
```

&nbsp;  
&nbsp;  

**Compare members and casual users**  

```{r, message = FALSE, warning = FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

&nbsp;  
&nbsp;  

**See the average ride time by each day for members vs casual users**  

```{r, message = FALSE, warning = FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

&nbsp;  
&nbsp;  

**Notice that the days of the week are out of order. Let's fix that.**  

```{r, message = FALSE, warning = FALSE}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

&nbsp;  
&nbsp;  

**Now, let's run the average ride time by each day for members vs casual users**  

```{r, message = FALSE, warning = FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

&nbsp;  
&nbsp;  

**Analyze ridership data by type and weekday**  

```{r, message = FALSE, warning = FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%   #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                   #groups by usertype and weekday
  summarise(number_of_rides = n()			#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		   # calculates the average duration
  arrange(member_casual, weekday)						# sorts
```

&nbsp;  
&nbsp;  

**Let's visualize the number of rides by rider type**  

```{r echo=TRUE, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
p <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "NUMBER OF RIDES by rider type", subtitle = "Cyclistic daily rentals", x = "Weekday", y = "Number of Rides", fill = "Rider Type")
ggplotly(p)
```




&nbsp;  
&nbsp; 

**Let's create a visualization for average trip duration**  

```{r echo=TRUE, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
q <- all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  labs(title = "AVERAGE TRIP DURATION by rider type", subtitle = "Cyclistic daily rentals", x = "Weekday", y = "Average Trip Duration", fill = "Rider Type")
ggplotly(q)
```
&nbsp;  
&nbsp;  

**Export summary files for further analysis**  
Create csv files that I will analyze in Excel and present findings.

```{r Export Summary Ride Length, message = FALSE, warning = FALSE}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "avg_ride_length.csv")
```


```{r Export Summary Number Of Rides, message = FALSE, warning = FALSE}
ride_counts <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n()) %>% 
    arrange(member_casual, weekday)
write.csv(ride_counts, file = "num_of_rides.csv")
```

&nbsp;  

### Key Findings  
  
  
Below are some key findings from the analysis:  
&nbsp;  
• Members ride bikes more often on weekdays (Monday-Friday), and less often on the weekend (Saturday and Sunday). Average number of rides by members on weekdays is 483587 and on weekends is 277962.  
&nbsp;  
• Casual riders rent bikes more often during the weekend (Saturday and Sunday), and less often on weekdays (Monday-Friday).  
Average number of rides by casual riders on weekend is 195418 and on weekdays 102269.  
&nbsp;  
• Average trip duration for members is 14 minutes 21 seconds and it's pretty consistent every day, with only a slight increase on the weekend.  
&nbsp;  
• Average trip duration for casual riders is 59 minutes 40 seconds, and it also does not fluctuate much during the week.  
&nbsp;    
• **Members use bikes much more often but take short trips**, most likely commuting to school or work.  
  **Casual riders rent bikes less often, however they use them for much longer trips**, probably sightseeing and relaxing.  

&nbsp;  

### Recommendations  

1. Introduce a weekend pass, priced below the annual pass.  
&nbsp;  
2. Emphasize longer ride duration allowed on the weekend pass in marketing material.  
&nbsp;  
3. It is difficult to propose precise plan pricing for the weekend pass without gathering more data, specifically regarding current annual membership pricing, as well as rental fees paid by casual riders at the moment. The suggested weekend pass would have to be priced attractively in order to convert casual riders, who only use bikes for pleasure, into Cyclistic members.  









